Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-07-27 doc/thesis/discussion.tex
Defence version of thesis with changes

\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
were sub-linear, ranging from 
    0.27
to 
    0.9.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. When used to analyze
population-level longitudinal partner count data, they found \gls{alpha} values
ranging from 0.26 to 0.62 depending on the gender and time period considered.
Our estimates of \gls{alpha} for the IDU/Romania was above this range under
both priors, as were the estimates for the MSM/Beijing data and the BC data
when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.}

@@

\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
IDU/Romania and IDU/Estonia, were 
    747 and 
    372.
The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
that 494 new \gls{HIV} infections were diagnosed among the studied population
(\glspl{IDU} in Bucharest) during the study period, and this value was
contained in our 95\% \gls{HPD} interval
    (136 -
    2379).
On the other hand, the IDU/Estonia data were collected from several locations
in the country. Using their reported incidence estimates, there were
approximately 2000 new infections in the country during the study period, a
value outside our 95\% \gls{HPD} interval
    (188 -
    834).
The authors indicated that their sample was ``convenience based''; it is
possible that other, harder-to-reach individuals had no contacts with those
sampled, reducing the apparent network size. In contrast, the IDU/Canada data
was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
population that has been intensively targeted by public health interventions.
Combined with the routine clinical genotyping performed in British Columbia for
all new \gls{HIV} infections, it is not unreasonable to suppose that the
sampled individuals (399) comprise the majority of \gls{HIV}-positive network
members. This value was the lower bound of our estimated 95\% \gls{HPD}
interval
    (399 -
    2862),
but it seems likely that the posterior mean of
    1481
is an overestimate.}

\add{The prevalence estimates for two \gls{MSM} populations in major Chinese
cities (MSM/Beijing and MSM/Shanghai) were 
    676 and
    1391.
The individuals sampled for the MSM/Beijing study were recruited by following a
prospective cohort of size 2000. 179 new infections were identified in the
cohort during the study period, which is much lower than the estimated
prevalence obtained with \software{netabc}. However, the authors did not claim
to have sampled the entire sexual network. In a nationwide survey,
\textcite{wu2013hiv} estimated that there were 24,198 \gls{MSM} living in
Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected, which is close to the
upper limit of the 95\% confidence interval of our estimate of \gls{I}
    (175 -
    1401).
The same survey estimated the size of the \gls{MSM} population in Shanghai
14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
This was within the 95\% confidence interval of the estimate obtained with
\software{netabc} for the MSM/Shanghai data
    (311 -
    2822).
It is worth noting that the authors of the IDU/Shanghai
study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
Shanghai, which is a significantly different estimate than that obtained by
\textcite{wu2013hiv}. The authors of the MSM/Taiwan
study~\autocite{kao2011surveillance} stated that there were 18,378
\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
    (268 -
    772),
suggesting that the entire \gls{MSM} population does not comprise a single
network. In fact, our estimate of \gls{I} for these data
    (524),
was closer to the number of \gls{MSM} enrolled in the study (301) than to the
epidemiologically estimated prevalence. The study focused exclusively on newly
diagnosed individuals. On the other hand, our estimated prevalence for the
MSM/USA data
    (482)
was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
individuals, which should have been a lower bound on \gls{I}, although the
value was contained in the 95\% \gls{HPD} interval
    (180 -
    1112).}

\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
populations in sub-Saharan Africa, where the dominant mode of transmission is
believed to be heterosexual. The HET/Uganda study population included
1434 seropositive individuals. Since the majority of inhabitants of the study
district were enrolled in the study, it is likely that this number represents
the majority of \gls{HIV} infected persons. Our estimate was slightly lower
    (939),
but the value 1434 was contained within the 95\% \gls{HPD} interval
    (226 -
    2661).
The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
sequencing effort, carried out on samples collected for other purposes between
1981 and 1989. Every individual in the study district was sampled; among these,
a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
lowest estimated \gls{I} value among the heterosexual datasets, 
the posterior mean
    (919),
was much higher than this value. Due to the thorough population-level
screening, it is most likely that the posterior mean represents an
overestimate, especially considering the observed upward bias of \gls{ABC}
estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
It is also possible that the true sexual network extended outside the study's
geographic area to include a larger number of infected individuals. The authors
of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
individuals in the study area. \gls{HIV} sequences were obtained from
approximately 70\% of these individuals. The estimated prevalence we obtained
was much higher 
    (5432),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
authors noted that the town where the study took place was located proximally
to the country's capital city, and suggested that frequent travel between the
two locations may have facilitated linking of their sexual networks. Thus, the
high estimate of \gls{I} we obtained for these data may include a larger
network component based in the capital city.}

>>>>

\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
were sub-linear, ranging from 
    0.27
to 
    0.81.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. When used to analyze
population-level longitudinal partner count data, they found \gls{alpha} values
ranging from 0.26 to 0.62 depending on the gender and time period considered.
Our estimates of \gls{alpha} for the IDU/Romania was above this range under
both priors, as were the estimates for the MSM/Beijing data and the BC data
when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV}-infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.}

@@

\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
IDU/Romania and IDU/Estonia, were 
    747 and 
    465.
The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
that 494 new \gls{HIV} infections were diagnosed among the studied population
(\glspl{IDU} in Bucharest) during the study period, and this value was
contained in our 95\% \gls{HPD} interval
    (136 -
    2379).
On the other hand, the IDU/Estonia data were collected from several locations
in the country. Using their reported incidence estimates, there were
approximately 2000 new infections in the country during the study period, a
value outside our 95\% \gls{HPD} interval
    (171 -
    1343).
The authors indicated that their sample was ``convenience based''; it is
possible that other, harder-to-reach individuals had no contacts with those
sampled, reducing the apparent network size. In contrast, the IDU/Canada data
was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
population that has been intensively targeted by public health interventions.
Combined with the routine clinical genotyping performed in British Columbia for
all new \gls{HIV} infections, it is not unreasonable to suppose that the
sampled individuals (399) comprise the majority of \gls{HIV}-positive network
members. This value was the lower bound of our estimated 95\% \gls{HPD}
interval
    (384 -
    3484),
but it seems likely that the posterior mean of
    1356
is an overestimate.}

\add{The prevalence estimates for two \gls{MSM} populations in major Chinese
cities (MSM/Beijing and MSM/Shanghai) were 
    676 and
    1391.
The individuals sampled for the MSM/Beijing study were recruited by following a
prospective cohort of size 2000. 179 new infections were identified in the
cohort during the study period, which is much lower than the estimated
prevalence obtained with \software{netabc}. However, the authors did not claim
to have sampled the entire sexual network. In a nationwide survey,
\textcite{wu2013hiv} estimated that there were 24,198 \gls{MSM} living in
Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected, which is close to the
upper limit of the 95\% confidence interval of our estimate of \gls{I}
    (175 -
    1401).
The same survey estimated the size of the \gls{MSM} population in Shanghai
14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
This was within the 95\% confidence interval of the estimate obtained with
\software{netabc} for the MSM/Shanghai data
    (311 -
    2822).
It is worth noting that the authors of the IDU/Shanghai
study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
Shanghai, which is a significantly different estimate than that obtained by
\textcite{wu2013hiv}. The authors of the MSM/Taiwan
study~\autocite{kao2011surveillance} stated that there were 18,378
\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
    (268 -
    607),
suggesting that the entire \gls{MSM} population does not comprise a single
network. In fact, our estimate of \gls{I} for these data
    (410),
was closer to the number of \gls{MSM} enrolled in the study (301) than to the
epidemiologically estimated prevalence. The study focused exclusively on newly
diagnosed individuals. On the other hand, our estimated prevalence for the
MSM/USA data
    (482)
was clearly an underestimate. The study enrolled almost 648 \gls{HIV}-infected 
individuals, which should have been a lower bound on \gls{I}, although the
value was contained in the 95\% \gls{HPD} interval
    (180 -
    1112).}

\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
populations in sub-Saharan Africa, where the dominant mode of transmission is
believed to be heterosexual. The HET/Uganda study population included
1434 seropositive individuals. Since the majority of inhabitants of the study
district were enrolled in the study, it is likely that this number represents
the majority of \gls{HIV}-infected persons. Our estimate was slightly lower
    (939),
but the value 1434 was contained within the 95\% \gls{HPD} interval
    (226 -
    2661).
The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
sequencing effort, carried out on samples collected for other purposes between
1981 and 1989. Every individual in the study district was sampled; among these,
a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
lowest estimated \gls{I} value among the heterosexual datasets, 
the posterior mean
    (724),
was much higher than this value. Due to the thorough population-level
screening, it is most likely that the posterior mean represents an
overestimate, especially considering the upward bias of \gls{ABC} estimates of
\gls{I} observed on simulated data (\cref{fig:abcpt2}). It is also possible
that the true sexual network extended outside the study's geographic area to
include a larger number of infected individuals. The authors of the
HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
estimated that there were 1731 \gls{HIV}-positive individuals in the study
area. \gls{HIV} sequences were obtained from approximately 70\% of these
individuals. The estimated prevalence we obtained was much higher 
    (5432),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
authors noted that the town where the study took place was located proximally
to the country's capital city, and suggested that frequent travel between the
two locations may have facilitated linking of their sexual networks. Thus, the
high estimate of \gls{I} we obtained for these data may include a larger
network component based in the capital city.}
